Abstractions for Multi-Sorted Substitutions

### 2023-09-18 paper/paper.tex
Update.

  Further repetition can be introduced by the choice of variable
  representation. For example, for de Bruijn indices, substitution is
  usually defined in terms of renamings to allow for structural
  induction.
  With a naive approach, this would again double the amount of
  substitution machinery, as all definitions and lemmas need to be
  made first for renamings and then again for substitutions.

@@

  Clearly, this situation is in need of automation.
  Our framework approaches this problem using a combination of
  abstractions and reflection in the context of extrinsically typed,
  intrinsically scoped syntax with de Bruijn indices.
  The user can write a language specification using regular Agda
  definitions (no generic encodings) and our framework derives
  definitions and lemmas for untyped substitution, and provides an
  abstraction to prove type preservation for all substitution operations
  using a single lemma. For System F, our framework allows to prove
  subject reduction with only a single handwritten lemma for type
  preservation.
  Compared to many other approaches, we do not derive substitutions
  for each of the variable sorts and syntactic categories, but instead
  use a novel approach for defining syntax, which directly supports
  substitutions that replace all variable sorts in parallel and can be
  applied to all syntactic categories.
  By further unifying renamings and substitutions, we gain the ability
  to talk abstractly about any kind of renaming or substitution that
  can occur in the formalization.
  Together with abstractions to unify renamings and substitutions,
  this allows us to create an abstraction that is general enough to
  prove type preservation for all sub

@@

  Frameworks for programming languages metatheory can be broadly
  categorized by representation of variable bindings, representation
  of typing and scoping, and requirements on the meta-language.

@@

  \emph{Autosubst}\cite{DBLP:conf/itp/SchaferTS15}
  is a Coq framework, which derives parallel substitution definitions and
  lemmas for languages from annotated Coq syntax definitions using
  extrinsic typing, extrinsic scoping, and de Bruijn indices.
  The framework is implemented in Coq's tactic language LTAC and comes
  with a decision procedure for all assumption-free, equational
  substitution-lemmas.
  The implementation of Autosubst deals with multiple variable sorts
  by generating multiple substitutions and corresponding interaction lemmas.
  Autosubst does not provide any machinery for type preservation and
  does not unify renaming and substitution.

@@

@@

>>>>

  Further repetition can occur due to the choice of variable
  representation. For example, for de Bruijn indices, substitution is
  usually defined in terms of renamings to allow for structural
  induction.
  With a naive approach, this would again double the amount of
  substitution machinery, as all definitions and lemmas need to be
  made first for renamings and then again for substitutions.

@@

  Clearly, this situation is in need of automation.
  Our framework approaches this problem using a combination of
  abstractions and reflection in the context of extrinsically typed,
  intrinsically scoped syntax with de Bruijn indices.
  The user can write a language specification using regular Agda
  definitions (no generic encodings) and our framework derives
  definitions and lemmas for untyped substitution, and provides an
  abstraction to prove type preservation for all substitution operations
  using a single lemma. For System F, our framework allows to prove
  subject reduction with only a single handwritten lemma for type
  preservation.

  Compared to many other approaches, we do not derive substitutions
  for each of the variable sorts and syntactic categories, but instead
  use a novel approach for defining syntax, which directly supports
  substitutions that replace all variable sorts in parallel and can be
  applied to all syntactic categories.
  By further unifying renamings and substitutions, we gain the ability
  to talk abstractly about any kind of renaming or substitution that
  can occur in the formalization.
  Having unified the untyped substitution machinery, we build further
  abstractions to similarly unify typing and type preservation lemmas.
  Our reflection algorithm instanciates those abstractions

  Together with abstractions to unify renamings and substitutions,
  this allows us to create an abstraction that is general enough to
  prove type preservation for all sub

@@

  Frameworks for programming language metatheory can be broadly
  categorized by representation of variable bindings and
  representation of typing and scoping.

@@

  \emph{Autosubst}\cite{DBLP:conf/itp/SchaferTS15}
  is a Coq framework, which derives parallel substitution definitions and
  lemmas for languages from annotated Coq syntax definitions using
  extrinsic typing, extrinsic scoping, and de Bruijn indices.
  The framework is implemented in Coq's tactic language LTAC and comes
  with a decision procedure for all assumption-free, equational
  substitution-lemmas.
  The implementation of Autosubst deals with multiple variable sorts
  by generating multiple substitutions and corresponding interaction lemmas.

@@

  Both works do not provide machinery for type preservation and
  do not unify renaming and substitution (although the latter is listed as
  future work). We believe it should be possible to translate their
  decision procedure from vectorized substitutions to the setting of
  our multi-sorted substitutions.

@@

  This work is orthogonal but compatible to our approach.
  We can instantiate the system using our syntax type system,
  but then the semantics abstraction would work with terms
  which are untyped with respect to the actual type system.
  We think it is worthwile to explore if their semantics abstraction
  can also be generalized to multi-sorted typing relations.
